---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data 

After loading the data using read.csv(), we transform the $date column of the data into class 'date' using as.date() for later use in creating the plots for analysis:

```{r}
unzip("activity.zip")
data <- read.csv("activity.csv")
data$date <- as.Date(as.character(data$date))
```

## What is mean total number of steps taken per day?

First, we will determine the # of steps taken on each observed day within the dataset:
```{r}
stepsbyday <- with(data,tapply(data$steps,data$date,sum, na.rm = T))
```

Next, we can create a histogram of the resulting data (i.e., the sum of steps taken per each observed day):
```{r}
hist(stepsbyday)
```

Lastly, we can determine the mean and median of the total number of steps taken per day by using the tapply and mean/median functions:
```{r}
with(data,tapply(data$steps,data$date,mean, na.rm = T))
```

```{r}
with(data,tapply(data$steps,data$date,median))
```

## What is the average daily activity pattern?
To visualize the average daily activity pattern, we can average the steps across all observed days for each 5-minute interval and graph the means in a time-series plot. First, we calculate the average steps taken within each 5-minute interval across all observed days by using the tapply function coupled with the mean function:
```{r, results = 'hide'}
z <- with(data,tapply(data$steps,data$interval,mean,na.rm = T))
```
Then, we create a plot using the names of the 'z' variable (aka. the 5-minute intervals) and the mean steps taken within each interval across all days ('z'):

```{r}
plot(names(z),z,type = "l")
```

To determine the 5-minute interval with the maximum mean number of steps, we can use the following code: 
```{r}
names(z[match(max(z),z)])
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
